Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-09-06 paper/paper.tex
Continue

Equipped with the above, we define the interpretation function \lstinline!&\Ist&! for set terms that interprets a set term with respect to a valuation function \lstinline!M :: 'a $\Rightarrow$ V! for variable.
\begin{lstlisting}
fun &\Ist& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  &\Ist& M $\emptyset$ = 0
| &\Ist& M (Var x) = M x
| &\Ist& M (Single s) = vset {&\Ist& M s}
| &\Ist& M (s $\unionS$ t) = &\Ist& M s $\sqcup$ &\Ist& M t 
| &\Ist& M (s $\interS$ t) = &\Ist& M s $\sqcap$ &\Ist& M t 
| &\Ist& M (s $\diffS$ t) = &\Ist& M s $-$ &\Ist& M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun &\Isa& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ bool
where
  &\Isa& M (s $\inS$ t) $\longleftrightarrow$ &\Ist& M s $\in$ elts (&\Ist& M t)
| &\Isa& M (s $\eqS$ t) $\longleftrightarrow$ &\Ist& M s = &\Ist& M t
\end{lstlisting}
For formulae, we write \lstinline!M $\models$ $\phi$! for the judgement that $\phi$ holds under the valuation function \lstinline!M!.
We omit the implementation of $\models$ for brevity.

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \textit{closed} if all of its branches are closed.

@@

\noindent The final step towards obtaining a proper model is to connect the realization function to the semantics as defined in Section~\ref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realize (Var x)) t = realize t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}
lemma coherence:
  assumes $\phi$ $\in$ set b
  shows ($\lambda$x. realize (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus since it gives us a model for every formula in an open and saturated branch.

\section{Soundness of the Calculus}
Two aspects determine the soundness of the calculus: a closed branch must be unsatisfiable and all expansion rules must maintain satisfiability.
We can express the former aspect in Isabelle as follows.
\begin{lstlisting}[belowskip=0pt]
lemma
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  Assume for contradiction that \lstinline!b! is closed but we have a model, i.e.\ it holds that \lstinline!$\forall$$\phi$ $\in$ set b. M $\models$ $\phi$!.
  We perform a case analysis on \lstinline!bclosed b!.
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  This leaves us with the case where \lstinline!b! contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}
  By assumption, each atom of the membership cycle holds under the model \lstinline!M!; in other words, we have
\begin{lstlisting}
  &\Ist& M t$_\text{i}$ $\in$ elts (&\Ist& M t$_\text{i + 1 mod k + 1}$)
\end{lstlisting}
  for each \lstinline!i $\in$ {0,$\ldots$,k}!.
  Therefore, the transitive closure of the membership relation \lstinline!{(x, y). x $\in$ elts y}! contains a cycle. 
  This, however, is a contradiction since the membership relation of \lstinline!V! is well-founded (see Section~\ref{sec:semantics}) and well-foundedness is closed under transitive closure.
\end{proof}

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by rule induction suffices to obtain the lemma below.
\begin{lstlisting}
lemma
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b'
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
The same would also hold for branching rules if it were not for the last rule that adds additional variables.
Those variables need to be assigned specific values as we will see in the proof below.
\begin{lstlisting}[belowskip=0pt]
lemma
  assumes &\fexpandsW{bs'}{b}&
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'.
          $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  Since the last branching rule is applied here, it holds that
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t],
         [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!, i.e.\ \lstinline!x $\notin$ vars b! and in particular \lstinline!x $\notin$ vars s! as well as \lstinline!x $\notin$ vars t!.
  Another consequence of applying the last rule is that \lstinline!s $\neqS$ t! is in \lstinline!b!, which, because \lstinline!M! is a model, implies that \lstinline!&\Ist& M s $\neq$ &\Ist& M t!.
  This inequality manifests itself through some \lstinline!y! with
\begin{lstlisting}
  &\phantom{$\lor$}& y $\in$ elts (&\Ist& M s) $\land$ y $\notin$ elts (&\Ist& M t)
  $\lor$ y $\in$ elts (&\Ist& M t) $\land$ y $\notin$ elts (&\Ist& M s).
\end{lstlisting}
  We now update the assignment \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that values of both \lstinline!s! and \lstinline!t! are unchanged under this update as \lstinline!x $\notin$ vars s! and \lstinline!x $\notin$ vars t!.
  Together with the our assumption about \lstinline!y!, we infer that 
\end{proof}

\section{Upper Bounding the Cardinality of a Branch}

\section{An Abstract Specification of the Decision Procedure}
\todo[inline]{Mention coherence lemma?}
\begin{figure*}
  \centering
\begin{lstlisting}
function mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b
  $\Longrightarrow$ mlss_proc_branch = mlss_proc_branch ((SOME b'. &\lexpands{b'}{b}& $\land$ set b $\subset$ set (b' @ b)) @ b)
| $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$
  $\Longrightarrow$ mlss_proc_branch b = ($\forall$b' $\in$ (SOME bs. &\fexpands{bs}{b}&). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True

definition mlss_proc :: 'a pset_fm $\Rightarrow$ bool where
  mlss_proc $\phi$ $\equiv$ mlss_proc_branch [$\phi$]
\end{lstlisting}
\caption{Definition of the functions \lstinline|mlss_proc_branch| and \lstinline|mlss_proc|.}
\end{figure*}

\subsection{Termination}
Maybe interesting: technical lemma that allows to obtain the point at which a parameter was introduced.
\subsection{Soundness}
Interesting: The model changes because auf new witnesses.
Maybe move this to the chapter on the calculus?
\todo[inline]{Show \lstinline!memberCycle! case? This requires \lstinline!member_seq!.}
\subsection{Lifting the Completeness of the Calculus}
\todo[inline]{Do you even lift?}

>>>>

Equipped with the above, we define the interpretation function \lstinline!&\Ist&! for set terms that interprets a set term with respect to a valuation function \lstinline!M :: 'a $\Rightarrow$ V! for variable.
\begin{lstlisting}
fun &\Ist& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  &\Ist& M $\emptyset$ = 0
| &\Ist& M (Var x) = M x
| &\Ist& M (Single s) = vset {&\Ist& M s}
| &\Ist& M (s $\unionS$ t) = &\Ist& M s $\sqcup$ &\Ist& M t 
| &\Ist& M (s $\interS$ t) = &\Ist& M s $\sqcap$ &\Ist& M t 
| &\Ist& M (s $\diffS$ t) = &\Ist& M s $-$ &\Ist& M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun &\Isa& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ bool
where
  &\Isa& M (s $\inS$ t) $\longleftrightarrow$ &\Ist& M s $\in$ elts (&\Ist& M t)
| &\Isa& M (s $\eqS$ t) $\longleftrightarrow$ &\Ist& M s = &\Ist& M t
\end{lstlisting}
For formulae, we write \lstinline!M $\models$ $\phi$! for the judgement that $\phi$ holds under the valuation function \lstinline!M!.
We omit the implementation of $\models$ for brevity.
\todo[inline]{Add implementation? Add satisfiability!}

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \textit{closed} if all of its branches are closed.
\todo[inline]{Move closed}

@@

\noindent The final step towards obtaining a proper model is to connect the realization function to the semantics as defined in Section~\ref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realize (Var x)) t = realize t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}
lemma coherence:
  assumes $\phi$ $\in$ set b
  shows ($\lambda$x. realize (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus as it gives us a model for every formula in an open and saturated branch.

\section{Soundness of the Calculus}
Two aspects determine the soundness of the calculus: a closed branch must be unsatisfiable and all expansion rules must maintain satisfiability.
We can express the former aspect in Isabelle as follows.
\begin{lstlisting}[belowskip=0pt]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  Assume for contradiction that \lstinline!b! is closed but we have a model, i.e.\ it holds that \lstinline!$\forall$$\phi$ $\in$ set b. M $\models$ $\phi$!.
  We perform a case analysis on \lstinline!bclosed b!.
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  This leaves us with the case where \lstinline!b! contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}
  By assumption, each atom of the membership cycle holds under the model \lstinline!M!; in other words, we have
\begin{lstlisting}
  &\Ist& M t$_\text{i}$ $\in$ elts (&\Ist& M t$_\text{i + 1 mod k + 1}$)
\end{lstlisting}
  for each \lstinline!i $\in$ {0,$\ldots$,k}!.
  Therefore, the transitive closure of the membership relation \lstinline!{(x, y). x $\in$ elts y}! contains a cycle. 
  This, however, is a contradiction since the membership relation of \lstinline!V! is well-founded (see Section~\ref{sec:semantics}) and well-foundedness is closed under transitive closure.
\end{proof}

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by rule induction suffices to obtain the lemma below.
\begin{lstlisting}
lemma lexpands_sound:
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b'
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
A similar argument would work for the branching rules if it were not for the last rule adding new variables.
Those variables need to be assigned specific values, hence modifying the model, as we will see in the proof below.
\begin{lstlisting}[belowskip=0pt]
lemma fexpands_sound:
  assumes &\fexpands{bs'}{b}&
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'.
          $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  We perform a case analysis on the derivation \lstinline!&\fexpands{bs'}{b}&!.
  As all cases but that of the last rule are straightforward, we only consider this last case where we have 
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t],
         [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!, i.e.\ \lstinline!x $\notin$ vars b! and in particular \lstinline!x $\notin$ vars s $\cup$ vars t!.
  Another consequence of applying the last rule is that \lstinline!s $\neqS$ t! is in \lstinline!b!, which, because \lstinline!M! is a model, implies that \lstinline!&\Ist& M s $\neq$ &\Ist& M t!.
  This inequality manifests itself through some \lstinline!y! with
\begin{lstlisting}
  &\phantom{$\lor$}& y $\in$ elts (&\Ist& M s) $\land$ y $\notin$ elts (&\Ist& M t)
  $\lor$ y $\in$ elts (&\Ist& M t) $\land$ y $\notin$ elts (&\Ist& M s).
\end{lstlisting}
  We now update the assignment \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that values of both \lstinline!s! and \lstinline!t! are unchanged under this update as \lstinline!x $\notin$ vars s! and \lstinline!x $\notin$ vars t!.
  Considering our assumption about \lstinline!y! and the fact that \lstinline!&\Ist& M' (Var x) = y!, we infer that
\begin{lstlisting}
  $\exists$b' $\in$ bs'. $\forall$$\psi$ $\in$ set b'. M' $\models$ $\psi$.
\end{lstlisting}
  In other words, due to the disjunction, it holds that \lstinline!M'! is a model for one \lstinline!b' $\in$ bs!.
  Thus, we are left with proving that \lstinline!M'! is a model for \lstinline!b!.
  This is indeed the case since the assumption \lstinline!M $\models$ $\psi$! is equivalent to \lstinline!M' $\models$ $\psi$! for all \lstinline!$\psi$ $\in$ set b! given that \lstinline!x! is fresh with respect to b.
\end{proof}

\section{Upper Bounding the Cardinality of a Branch}
Soundness and completeness of the calculus are the building blocks with which we establish partial correctness of Algorithm~\ref{fig:alg}.
Consequently, we still need to give a termination proof of Algorithm~\ref{fig:alg} to obtain total correctness of the algorithm.
The core of our argument is that the number of distinct formulae in a branch that can be derived with the expansion rules is finite.
In particular, we claim that the number of distinct formulae has the following upper bound.
\begin{lstlisting}
  &\phantom{$\leq$}& card (set b)
  &\textover{$=$}{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$} +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& 16 * (card (subterms (last b)))$^4$
\end{lstlisting}
We justify both inequalities separately, starting with the first one whose justification hinges on an invariant:
Its justification hinges on the invariant that the only new subformulae introduced by the expansion rules introduce

\section{An Abstract Specification of the Decision Procedure}
\begin{figure*}
  \centering
\begin{lstlisting}
function mlss_proc_branch :: 'a branch $\Rightarrow$ bool where
  $\neg$ lin_sat b
  $\Longrightarrow$ mlss_proc_branch = mlss_proc_branch ((SOME b'. &\lexpands{b'}{b}& $\land$ set b $\subset$ set (b' @ b)) @ b)
| $\llbracket$ $\neg$ sat b; bopen b; lin_sat b $\rrbracket$
  $\Longrightarrow$ mlss_proc_branch b = ($\forall$b' $\in$ (SOME bs. &\fexpands{bs}{b}&). mlss_proc_branch (b' @ b))
| $\llbracket$ lin_sat b; sat b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = bclosed b
| $\llbracket$ lin_sat b; bclosed b $\rrbracket$ $\Longrightarrow$ mlss_proc_branch b = True

definition mlss_proc :: 'a pset_fm $\Rightarrow$ bool where
  mlss_proc $\phi$ $\equiv$ mlss_proc_branch [$\phi$]
\end{lstlisting}
\caption{Definition of the functions \lstinline|mlss_proc_branch| and \lstinline|mlss_proc|.\label{fig:alg}}
\end{figure*}
